---
title: "HWA2.Rmd"
output: html_document
---
---
title: "R intermediate"
author: "Dan McGlinn"
date: "January 15, 2016"
output: html_document
---

Examine the following for loop, and then complete the exercises

```{r}
data(iris)
head(iris)

sp_ids <- unique(iris$Species)

output <- matrix(0, nrow=length(sp_ids), ncol=ncol(iris)-1)

rownames(output) <- sp_ids
colnames(output) <- names(iris[ , -ncol(iris)])

for(i in seq_along(sp_ids)) {
    iris_sp = subset(iris, subset=Species == sp_ids[i], select=-Species)
    for(j in 1:(ncol(iris_sp))) {
        x <- 0
        n_samp <- 0
            for(k in 1:nrow(iris_sp)) {
                x <- x + iris_sp[k, j]
                n <- n_samp + 1
            }
            output[i, j] = x / n
        }
}
output

x

```
##Excercises
###Iris loops

```{r}
# what is the for loop doing? Start with seq_along(sp_ids)

seq_along(sp_ids)
# gives 1, 2, 3

subset(iris, subset=Species == sp_ids[i], select=-Species)
# gives iris data (sepal.length, sepal.width, petal.length, petal.width)
# gets rid of "species" columns

for(j in 1:(ncol(iris_sp))) {
        x <- 0
        y <- 0
}

i
# i gives the value 3

j
# j gives the value 4

```


1. Describe the values stored in the object `output`. In other words what did the 
loops create?
The values stored in the object `output` is a table comparing sepal petals, length, and width for the setosa, versicolor, and virgincia. 

2. Describe using pseudo-code how `output` was calculated, for example,
```{r, eval=FALSE}
Loop from 1 to length of species identities
   Take a subset of iris data 
   Loop from 1 to number of columns of the iris data
      If ... occurs then do ...  
      
Loop from 1 to length of species ids
   Use Iris data exluding species column
      Loop from 1 to the last row of iris species
        Define x and n_samp
          Add all values of x
          Add the number of arguments used to find x
          Divide the sum of all x by the number of arguments
Print output
          
```

3. The variables in the loop were named so as to be vague. How can the objects
`output`, `x`, and `y` could be renamed such that it is clearer what is occurring
in the loop.

Output would be more clearly named as "average"

'x' would be more clearly named col_sum

'y' would be more clearly names n_samp, as it represents the number of columns in samples

4. It is possible to accomplish the same task using fewer lines of code? Please suggest one other way to calculate `output` that decreases the number of loops
by 1. 

```{r}
# loop through species names
for(i in seq_along(sp_ids)) {
    # subset data for species species
    iris_sp = subset(iris, subset=Species == sp_ids[i], select=-Species)
    
    for(j in 1:(ncol(iris_sp))) {
            output[i, j] <- mean(iris_sp[ , j]) 
    }
}
output

```

###Sum of a sequence

5. You have a vector `x` with the numbers 1:10. Write a
for loop that will produce a vector `y` that contains the sum of `x` up to that
index of `x`. So for example the elements of `x` are 1, 2, 3, and so on and the 
elements of `y` would be 1, 3, 6, and so on. 

```{r}

y <- NULL
x <- c(1:10) # defines values of vector x
for (i in x) { # establishes parameters in loop 
  y[i] = sum(x[1:i]) # defines function of y and summation for vector output
}
y # prints 'y' as output

```

6. Modify your for loop so that if the sum is greater than 10 the value of `y`
is set to NA

```{r}

y <- NULL
x <- c(1:10)
for (i in x) {
  y[i] = sum(x[1:i]) # same function as defined in q5
  if (y[i]<=10) { # establishes conditions of less than/equal to 10
    print(y) # prints values for y less than 10
  } else if (y[i]>10) { # establishes greater than/equal to 10 condition
    print (NA) # directions for what to do if above 10
  }
} 
y[i] <- NA

```

7. Place your for loop into a function that accepts as its argument any vector
of arbitrary length and it will return `y`.

```{r}

vector_add <- function(x,y) {
    output <- NULL
    x <- (1:n) # define variables so they aren't taken from r global
    y <- NULL
    for (i in x) { 
  y[i] = sum(x[1:i])
     }
     vector_add
}

print(vector_add)


```
