---
layout: page
---

## Univariate Assignment

Read in tree data

```{r}
# read in directly from website: 
trees <- read.csv('https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/data/treedata_subset.csv')
# or download and import locally
treedata_subset <- read.csv("~/Desktop/treedata_subset.csv")

```

1\. Carry out an exploratory analysis using the tree dataset. Metadata for the
tree study can be found [here](../data/tree_metadata.txt). Specifically, I would
like you to develop and compare models for species cover for a habitat
generalist [*Acer rubrum* (Red
maple)](http://www.durhamtownship.com/blog-archives/pix/November1407.jpg) and a
habitat specialist [*Abies fraseri* (Frasier
fir)](https://upload.wikimedia.org/wikipedia/commons/d/d0/Abies_fraseri_Mitchell.jpg).
Because this dataset includes both continuous and discrete explanatory variables
use the function `Anova` in the packages `car` as such

```{r}
str(trees)
# cover is integer (ordinal numbers), species is a factor (name)

```

This will estimate partial effect sizes, variance explained, and p-values for 
each explanatory variable included in the model. 

Compare the p-values you observe using the function `Anova` to those generated
using `summary`. 

For each species address the following additional questions:

* how well does the exploratory model appear to explain cover?
The exploratory model explains about half of the total variance of the model (when comparing residuals). 

* which explanatory variables are the most important?
Species, elevation, beers, TCI, and streamdist (in this order) are the most important explanatory variables when acer and abies cover are grouped. Each of these values were significant. Species was the most important factor by far, and elevation followed. When acer is ran alone, the most important values are elev, tci, and beers. When abies is ran alone, the most important variables are elev, stream dist, then tci. 

* do model diagnostics indicate any problems with violations of OLS assumptions?
The variable cover looks is affecting the residuals plot because it is an ordinal variable, meaning that it is a group of numbers that is representing different 'bins' of data, not a continuous variable as the model assumptions specify. Additionally, the data is not normally distributed. The assumptions of normality and equal variance of the residuals are not met. 

* are you able to explain variance in one species better than another, 
  why might this be the case?
In the quadrats sampled, there was substantially more red maple than fraiser fir (lots of 0's reported for species cover in abies; out of 734 entries, there are about 26 entries noting the presence of fraiser fir). This would affect the variance of this data set compared to the red maple data set, as fewer observations would skew the data. 

Prior to addressing the above questions you will want to restructure and 
subset the data using the following R code: 

```{r Dan's code for function, annotated}  
# we wish to model species cover across all sampled plots
# create site x sp matrix for two species 
sp_cov = with(trees, tapply(cover, list(plotID, spcode), 
                           function(x) round(mean(x))))
# applying a function
sp_cov = ifelse(is.na(sp_cov), 0, sp_cov)
sp_cov = data.frame(plotID = row.names(sp_cov), sp_cov)
# create environmental matrix
cols_to_select = c('elev', 'tci', 'streamdist', 'disturb', 'beers')
env = aggregate(trees[ , cols_to_select], by = list(trees$plotID), 
                function(x) x[1])
names(env)[1] = 'plotID'
# merge species and environmental matrices
site_dat = merge(sp_cov, env, by='plotID')
# subset species of interest
abies = site_dat[ , c('ABIEFRA', cols_to_select)] 
acer  = site_dat[ , c('ACERRUB', cols_to_select)]
# subset for each species of interest
names(abies)[1] = 'cover' 
names(acer)[1] = 'cover'
# renaming these columns to cover
# new variable created called sp_cov
```


```{r making plots}

plot(abies)
# seems to be a relationship between TCI and stream dist

plot(acer)
# seems to be same relationship between TCI and stream dist

abies$species = "abies" # making a new column
acer$species = "acer"

spc_interest = rbind(abies, acer)
View(spc_interest)

boxplot(cover~species, data = spc_interest)
# shows box plot: box plot is kind of not useful because abies data is grouped together near the bottom

```

```{r model making}

library(car) # referencing this ensures that when you look back you'll know which package is being used
species_cover.aov = aov(cover~species, data = spc_interest) # makes anova model
# models data independently of species

Anova(species_cover.aov, type = 3) # gives results of anova model

summary(species_cover.aov) # gives summary of model fit

spc_cov_fac.aov = aov(cover~species+elev+tci+streamdist+disturb+beers, data = spc_interest) # makes anova model with more factors and their interactions

Anova(spc_cov_fac.aov, type = 3)

summary(spc_cov_fac.aov)
# way too complicated of a model; need to simplify, too many interactions to analyze

step(spc_cov_fac.aov) # gives insight as to how to reduce complexity of model
# disturbance is not very insightful for model analysis/significance 
# compares AIC values

cover_best.aov = aov(cover~species+elev+tci+streamdist+beers, data = spc_interest)
summary (cover_best.aov)
Anova(cover_best.aov, type = 3)

cover_acer.aov = aov(cover~elev+tci+streamdist+beers, data = acer)
summary (cover_acer.aov)
Anova(cover_acer.aov, type = 3)
# gives information specific to acer

step(cover_acer.aov)
# elevation is most important factor affecting acer; streamdist is least important

cover_acer_best.aov = aov(cover~elev+tci+beers, data = acer)
summary(cover_acer_best.aov)
Anova(cover_acer_best.aov)


cover_abies.aov = aov(cover~elev+tci+streamdist+beers, data = abies)
summary (cover_abies.aov)
Anova(cover_abies.aov, type = 3)
# gives information specific to abies

step(cover_abies.aov)
# elev is most important (changes aic most), then streamdist, then tci 
# when I run this, I get negative AIC: is this normal?

```

```{r checking model}

par(mfrow = c(2,2)) # orients output to 2 x 2 matrix of graphs for comparison
plot(cover_best.aov) # tests for assumptions

```

2\. You may have noticed that the variable cover is defined as 
positive integers between 1 and 10. and is therefore better treated
as a discrete rather than continuous variable. 
Re-examine your solutions to the question above but from the
perspective of a General Linear Model (GLM) with a Poisson error term
(rather than a Gaussian one as in OLS). 
The Poisson distribution generates integers 0 to positive infinity so this may provide a good first approximation. 
Your new model calls will look as follows:

```{r models with continuous variables} 

acer_poi = glm(cover~elev+tci+streamdist+beers, data = acer, 
           family='poisson')
summary (acer_poi) 

# model of acer data with poisson
# most significant variables: elevation, tci, beers

step(acer_poi)
# most altered AICs: elev, then tci, then beers; stream dist alters AIC least

abies_poi = glm(cover~elev+tci+streamdist+beers, data = abies, family='poisson')
summary (abies_poi)
# model of abies data with poisson
# most significant relationships: elev, then tci, then stream dist

step(abies_poi)
# most altered AICs: elev (by a lot), then tci and stream dist; beers alters aic least

```


```{r test poisson assumptions}
par(mfrow = c(2,2)) # orients output to 2 x 2 matrix of graphs for comparison
plot(acer_poi)
# More confirmation to normal distribution in this plot
# still a lot of data points all over graphs for scale-location and residuals

par(mfrow = c(2,2))
plot(abies_poi)
# this model conforms a lot better than earlier models

```
For assessing the degree of variation explained you can use a 
pseudo-R-squared statistic (note this is just one of many possible)

```{r}
pseudo_r2 = function(glm_mod) {
                1 -  glm_mod$deviance / glm_mod$null.deviance
            }
```

Compare your qualatitive assessment of which variables were most important in each model. 
For acer: most significant variables: elevation, tci, beers
  For acer in other model: elevation, tci, and beers
For abies: most significant relationships: elev, then tci, then stream dist
  For abies in other model: elevation, stream dist, tci. The order of importance was changed for this model

Does it appear that changing the error distribution changed the results much? In what ways? 
Changing the error distribution  affected the results of the two models (the importance of a few variables were altered between the preliminary model and the poisson model), but the poisson model fit the assumptions of the normality and residuals better. 

3\. Provide a plain English summary (i.e., no statistics) of what you have
found and what conclusions we can take away from your analysis?
(Based on the poisson model)
The coverage of two species of trees in the smoky mountains, the fraiser fir (abies) and the red maple (acer), were analyzed and their coverage was compared to the variables elevation, water potential (tci), their distance from the nearest stream, and their transformed slope/heat load index (beers). Using general linear testing assuming a poisson error distribution, it was found that the variables that most affected acer coverage were elevation, tci, and beers. The variables that most affected abies coverage were elevation, tci, and stream distance. 
These variables were found to have a significant relationship with the amount of tree coverage present, which was represented by numbers on a scale of 1-10 and had different percentage rankings to 'bin' coverage. Both species' cover were found to be most affected by elevation. 

4\. (optional) Examine the behavior of the function `stepAIC()` using the 
exploratory models developed above. This is a very simple and not very
robust machine learning stepwise algorithm that uses AIC to select a 
best model. By default it does a backward selection routine. 

5\. (optional) Develop a model for the number of species in each site 
(i.e., unique plotID). This variable will also be discrete so the Poisson
may be a good starting approximation. Side note: the Poisson
distribution converges asymptotically on the Gaussian distribution as the 
mean of the distribution increases. Thus Poisson regression does not differ
much from traditional OLS when means are large. 