---
title: "HWA4.Rmd"
output: html_document
---
---
layout: page
title: multivariate models
---

For this assignment will be analyzing data on the Vegetation
and Environment in Dutch Dune Meadows.

To import the data and read the metadata run the following:

```{r}
library(vegan)
data(dune)
data(dune.env)
?dune

dune.env
```

```{r}
dune_mds <- metaMDS(dune) # assigns metadata dune for the rest of analysis
```

1. Conduct an indirect ordination on the dune plant community. Specifically,
visually examine a NMDS plot using the bray-curtis distance metric. Below is 
some code to help you develop a potential plot that emphasizes the role of the
environmental variable "Moisture". Describe how you interpret the
graphic. What is the goal of creating such a plot? Does this analysis suggest
any interesting findings with respect to the dune vegetation?

    ```{r}
    plot(dune_mds, type='n')
    text(dune_mds, 'sp', cex=.5)
    # generate vector of colors 
    color_vect = rev(terrain.colors(6))[-1]
    points(dune_mds, 'sites', pch=19, 
           col=color_vect[dune.env$Moisture])
    legend('topright', paste("Moisture =", 1:5, sep=''), 
           col=color_vect, pch=19)
    
# moisture 4 are congregating to the right of 0, moisture 1 are to the left of 0
    ```
```{r}
# This graphic shows a grouping of the category "Moisture 4" to the right of 0 along the x axis, as well as a grouping of "Moisture 1" to the left of 0 along the x axis. There is some separation aspect along the Y axis, although at this point in analysis there is no discernible trend (further analyses may provide more insight). The purpose of producing this plot is to analyze how all factors are interacting while overlaid with the variable moisture; instead of using preliminary hypothesis testing, this method looks at many types of relationships at once. 
# This graphic suggests interesting findings in regards to how different species are interacting with the Moisture 1 and Moisture 4 variables. 
```

2. Carry out a direct ordination using CCA in order to test any potential 
hypotheses that you developed after examining the MDS plot. Specifically,
carry out a test of the entire model (i.e., including all constrained axes)
and also carry out tests at the scale of individual explanatory variables
you included in your model if you included more than one variable. Plot your 
results.
```{r}
dune.env$Moisture <- as.numeric(dune.env$Moisture)
dune.env$Manure <- as.numeric(dune.env$Manure)

cca_dune_all = cca(dune~Moisture+A1+Management, data=dune.env)
cca_dune_all
plot(cca_dune_all)
plot(cca_dune_all, type='n', scaling=1)
orditorp(cca_dune_all, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune_all, display='cn', col='red')
anova(cca_dune, by='margin', permutations = 999)

# variables use and manure are insignificant for this model

cca_dune_alls = cca(dune~Moisture+A1+Management+Manure+Use, data=dune.env)
cca_dune_alls
plot(cca_dune_alls)
plot(cca_dune_alls, type='n', scaling=1)
orditorp(cca_dune_alls, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune_alls, display='cn', col='red')
anova(cca_dune_alls, by='margin', permutations = 999)

cca_dune_A1 = cca(dune~A1, data=dune.env)
cca_dune_A1
plot(cca_dune_A1) 
anova(cca_dune_A1, permutations = 999) # a1 is significant

cca_dune_Man = cca(dune~Management, data=dune.env)
cca_dune_Man
plot(cca_dune_Man)
anova(cca_dune_Man, permutations = 999) # management is significant


cca_dune = cca(dune~Moisture, data=dune.env)
# run cca with moisture as an effect of dune
cca_dune # print cca
plot(cca_dune)


```

3. Do your two analyses agree with one another or complement one another or do
these two analyses seem to be suggesting different take home messages? Which
analysis do you find to be more useful?
```{r}
# Yes, my analyses do complement one another. Moisture appeared to significantly affect dune data in the NMDS plot from question 1, and anova testing and cca plots in question 2 confirm that this variable was significant. 
# CCA plots and anova testing from question 2 were more specific metrics for analyzing data. This suggested different findings than the NMDS because it gave more insight into each specific variable rather than a generalized approach.
# I found the anovas to be more useful because they gave me specific significance values. These were important for discerning relationships and it provides more insight about individual variables. 

```